Appendix B: System size expansion of the master equation for demographic noise

### 2018-03-05 appendixB/appendixB.Rmd
fix typos

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: ../paper/refs.bib
output: 
  rticles::elsevier_article:
    toc: true

preamble: "\\newcommand{\ud}{\\mathrm{d}}"
---

@@

$$
\begin{equation}
\frac{\mathrm{d}}{\mathrm{d}t} P(x,t) = \int w(x|y) P(y,t) - w(y|x)P(x,t) \mathrm{d} x
\end{equation}
$$

@@

\begin{align}
\intertext{Take as the definition}
\mathbb E^k f(n) &= f(n+k) \nonumber \\
\intertext{Then}
\mathbb E^k n &=  n+k \nonumber \\
\mathbb E^k n^2 &= \mathbb E^k n n = (n+k)(n+k) = n^2+2kn + k^2 \nonumber \\
\intertext{This suggests we can approximate the step operator by a Taylor series}
\mathbb E^k &= 1+ k\frac{\partial}{\partial n} + \frac{k^2}{2} \frac{\partial^2}{\partial n^2} + \ldots \label{En}
\end{align}
To change variables, recall the chain rule,
\begin{align}
&\frac{\partial}{\partial \xi} f(n(\xi)) =  \frac{\partial n}{\partial \xi} \frac{\partial}{\partial n} f(n(\xi)) \nonumber \\
\intertext{hence} 
&\frac{\partial}{\partial n} =  \left(\frac{\partial n}{\partial \xi}\right)^{-1} \frac{\partial}{\partial \xi} \nonumber 
\end{align}
Making this substitution to \eqref{En} we find:

@@

Also by the chain rule, we have
\begin{align*}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\mathrm{d} \xi}{\mathrm{d} t} 
\end{align*}
Substituting in \eqref{dxidt}, we find:

@@

$$
\begin{equation}
a_i = \int r^i \Phi(r) \mathrm{d} r
\end{equation}
$$
The first jump moment, $i == 1$ the two possible steps $r$ are $\pm 1$, and thus $a_1$ is sum of the two possible one-step transitions $+1 \cdot b(x) + -1 \cdot d(x)$.  The second jump moment the $r$ term is raised to the second power, making both contributions positive such that $a_2(x) = b(x) + d(x)$

@@

$$\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)$$
though the partition into birth and death rates is not explicit.  Different ways of dividing this equation between birth and death can therefore create different fluctuation patterns, even as the macroscopic average remains unchanged.  For instance, taking $b = rn$ and $d = rn^2/K$ and  we can calculate the fluctuations of the logistic equation around its equilibrium $n^* = K$ as:

@@

$$
\begin{align*}
\mathrm{d} y &= - \beta y \phantom \cdot \mathrm{d} t + \sigma_e \mathrm{d}W
\end{align*}
$$
which corresponds to mean-zero Gaussian noise with auto-correlation $\beta$, where the white-noise limit corresponds to $\beta \to 0$. Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $, while the fluctuation dynamics are now given by three coupled differential equations:

$$
\begin{align}
\frac{\mathrm{d} \langle \xi^2 \rangle}{\mathrm{d} t} &= 2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial y} \langle \xi \eta \rangle + a_2 \\
\frac{\mathrm{d} \langle \xi \eta \rangle}{\mathrm{d} t} &= \left( \frac{\partial a_1}{\partial x} - \beta \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial y} \langle \eta^2 \rangle \\
\frac{\mathrm{d} \langle \eta^2 \rangle}{\mathrm{d} t} &= -2 \beta \langle \eta^2 \rangle + \gamma_e^2
\end{align}

$$

@@

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{\frac{\left[\partial_{y} a_1\right] ^2}{\partial_x a_1 - \beta} \frac{\langle\eta^2\rangle}{2\partial_x a_1}}_{\textrm{environmental contribution}}
$$

@@

$$\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},$$

@@

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} +  \underbrace{\frac{1}{2} \left[\frac{\partial_{y} a_1}{ \partial_x a_1}\right] ^2}_{\textrm{environmental coupling}} \langle\eta^2\rangle
$$
Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:

$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2$$

@@

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{- \frac{\left[\partial_{y} a_1\right] ^2}{2\partial_x a_1 \beta}}_{\textrm{environmental coupling}}  \langle\eta^2\rangle
$$
(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).  In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher autocorelation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.

>>>>

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: "../paper/refs.bib"
output: 
  rticles::elsevier_article:
    toc: true
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@@

\begin{equation}
\frac{\mathrm{d}}{\mathrm{d}t} P(x,t) = \int w(x|y) P(y,t) - w(y|x)P(x,t) \mathrm{d} x
\end{equation}

@@

\begin{align}
\intertext{Take as the definition}
\mathbb E^k f(n) &= f(n+k) \nonumber \\
\intertext{Then}
\mathbb E^k n &=  n+k \nonumber \\
\mathbb E^k n^2 &= \mathbb E^k n n = (n+k)(n+k) = n^2+2kn + k^2 \nonumber \\
\intertext{This suggests we can approximate the step operator by a Taylor series}
\mathbb E^k &= 1+ k\frac{\partial}{\partial n} + \frac{k^2}{2} \frac{\partial^2}{\partial n^2} + \ldots \label{En}
\end{align}

To change variables, recall the chain rule,

\begin{align}
&\frac{\partial}{\partial \xi} f(n(\xi)) =  \frac{\partial n}{\partial \xi} \frac{\partial}{\partial n} f(n(\xi)) \nonumber \\
\intertext{hence} 
&\frac{\partial}{\partial n} =  \left(\frac{\partial n}{\partial \xi}\right)^{-1} \frac{\partial}{\partial \xi} \nonumber 
\end{align}

Making this substitution to \eqref{En} we find:

@@

Also by the chain rule, we have

\begin{align*}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\mathrm{d} \xi}{\mathrm{d} t} 
\end{align*}
Substituting in \eqref{dxidt}, we find:

@@

\begin{equation}
a_i = \int r^i \Phi(r) \mathrm{d} r
\end{equation}

The first jump moment, $i == 1$ the two possible steps $r$ are $\pm 1$, and thus $a_1$ is sum of the two possible one-step transitions $+1 \cdot b(x) + -1 \cdot d(x)$.  The second jump moment the $r$ term is raised to the second power, making both contributions positive such that $a_2(x) = b(x) + d(x)$

@@

\begin{equation*}
\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)
\end{equation*}

though the partition into birth and death rates is not explicit.  Different ways of dividing this equation between birth and death can therefore create different fluctuation patterns, even as the macroscopic average remains unchanged.  For instance, taking $b = rn$ and $d = rn^2/K$ and  we can calculate the fluctuations of the logistic equation around its equilibrium $n^* = K$ as:

@@

\begin{align*}
\mathrm{d} y &= - \beta y \phantom \cdot \mathrm{d} t + \sigma_e \mathrm{d}W
\end{align*}

which corresponds to mean-zero Gaussian noise with auto-correlation $\beta$, where the white-noise limit corresponds to $\beta \to 0$. Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\tfrac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y)$, while the fluctuation dynamics are now given by three coupled differential equations:

\begin{align}
\frac{\mathrm{d} \langle \xi^2 \rangle}{\mathrm{d} t} &= 2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial y} \langle \xi \eta \rangle + a_2 \\
\frac{\mathrm{d} \langle \xi \eta \rangle}{\mathrm{d} t} &= \left( \frac{\partial a_1}{\partial x} - \beta \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial y} \langle \eta^2 \rangle \\
\frac{\mathrm{d} \langle \eta^2 \rangle}{\mathrm{d} t} &= -2 \beta \langle \eta^2 \rangle + \gamma_e^2
\end{align}

@@

\begin{align}
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{\frac{\left[\partial_{y} a_1\right] ^2}{\partial_x a_1 - \beta} \frac{\langle\eta^2\rangle}{2\partial_x a_1}}_{\textrm{environmental contribution}}
\end{align}

@@

\begin{equation*}\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},\end{equation*}

@@

\begin{equation}
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} +  \underbrace{\frac{1}{2} \left[\frac{\partial_{y} a_1}{ \partial_x a_1}\right] ^2}_{\textrm{environmental coupling}} \langle\eta^2\rangle
\end{equation}

Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:

\begin{equation}
\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2
\end{equation}

@@

\begin{equation}
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{- \frac{\left[\partial_{y} a_1\right] ^2}{2\partial_x a_1 \beta}}_{\textrm{environmental coupling}}  \langle\eta^2\rangle
\end{equation}

(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).  In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher autocorelation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.
